### 2021-03-21 manuscript/v5_submission_revision/manuscript_tex4submission.tex
update manuscript for submission

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Causal analysis of COVID-19 observational data in German districts
reveals effects of mobility, awareness, and temperature} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Edgar Steiger\textsuperscript{1*},
Tobias Mußgnug\textsuperscript{1},
Lars Eric Kroll\textsuperscript{1}%,
%Name4 Surname\textsuperscript{2},
%Name5 Surname\textsuperscript{2\ddag},
%Name6 Surname\textsuperscript{2\ddag},
%Name7 Surname\textsuperscript{1,2,3*},
%with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Central Research Institute of Ambulatory Health Care in Germany (Zi),
Salzufer 8, D-10587 Berlin, Germany
\\
%\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
%\\
%\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
%\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Mobility, awareness, and weather are suspected to be causal drivers for new cases of COVID-19 infection. Correcting for possible confounders, we estimated their causal effects on reported case numbers. To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the aforementioned determinants on new reported cases of COVID-19. Based on this, we computed valid adjustment sets of the possible confounding factors. We collected data for Germany from publicly available sources (e.g.~Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression. Our analysis revealed favorable causal effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases. Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for ``corona'' in Google, and higher rainfall, leading to an increase in new COVID-19 cases. This comprehensive causal analysis of a variety of determinants affecting COVID-19 progression gives evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.

@@

Our objective is to provide estimates of the causal effects of the main
drivers of the pandemic with reduced bias. We conducted a scoping review
of the available studies regarding signaling pathways and determinants
of the spread of SARS-CoV-2 infections and the reported new COVID-19
cases. Then we integrated the current findings into a directed acyclic
graph for the progress of the pandemic at the regional level. Using the
resulting model and the do-calculus we found identifiable effects
without blocked causal paths whose effects can be analyzed with
observational data \cite{Pearl_2014}. We used regional time series data
of all 401 German districts from various publicly available
sources to analyze these questions on a regional level. Germany is a
good choice in this regard, because it has ample data on contributing
factors on the regional level and has had high testing and treatment
capacities from early on in the pandemic.

@@

When investigating causal determinants of SARS-CoV-2 infections, a
number of confounders have to be considered. Well-known risk factors for
SARS-CoV-2 as well as for other infections are demographic factors such
as age, gender, socio-economic status (SES), population density, and
foreign citizenship/ethnicity
\cite{de2020risk, Dragano2020.06.17.20133918, jhucovid19db2020}. In
Germany along with other countries (i.e.~Brazil, USA, or the UK),
populist parties or politicians and their electorate tend to be more
sceptical about effects of containment measures than the other part of
the electorate \cite{dohle_wingen_schreiber_2020, engle_staying_2020}.
Therefore we considered both ``right-wing populist party votes'' and
``voter turnout'' as possible confounders. Public health interventions
were also taken into account (contact restrictions, school closures
etc.), as their implementation showed strong correlations with
controlling the spread of SARS-CoV-2
\cite{cowling2020impact, juni_impact_2020, lai_effect_2020}. To avoid
bias due to reporting delay of case numbers we had to include weekday
and German holidays. We included some unobserved variables in our DAG
(e.g. ``Herd immunity''), too. Please note that ``Exposure to
SARS-CoV-2'' is itself an unobserved variable: German case numbers are
reported with delay after date of exposure and symptom onset.
\emph{Exposure to the virus} should not be confused with the formal
\emph{exposure variables} of the DAG (mobility, awareness, temperature).

@@

We collected and aggregated data on reported COVID-19 cases, regional
socio-demographic factors, weather, and general mobility on district and
state level in Germany for the period of 15 February 2020 to 8 July
2020. Our observation period for the outcome consisted of all dates from
23 February 2020 to 8 July 2020 (\(T=137\)), since we used a lag of 8
days for all confounders. We did not exclude any states or districts
(\(K=401\)). We analyzed the daily reported number of new cases as
outcome (\(K\cdot T=54{\ }937\) observations). The set of possible
predictors was derived from our causal DAG (see Table~\ref{tab:model-vars} and Fig~\ref{fig:dag-covid-19}). Due to
modelling and data limitations, some of the predictors were unobserved
or were modelled as a construct consisting of several variables. For our
causal analysis, we computed adjustment sets in three different
scenarios for separate exposures within the DAG: i) mobility of
population, ii) awareness of COVID-19 (i.e.~Google searches for
``corona''), iii) weather (i.e.~temperature).

@@

To assess the mobility of the German population, we used data publicly
available on German state level from Google \cite{google_mobility}.
Measurements are daily relative changes of mobility in percent compared
to the period of 3 January 2020 to 6 February 2020. Missing values
(\(25\) out of \(13{\ }488\)) were imputed with value \(0\) and the
state level measurements were passed onto districts within the
corresponding state. Google mobility data was available for six
different sectors of daily life (``retail and recreation'', ``grocery
and pharmacy'', ``parks'', ``transit stations'', ``workplaces'',
``residential'') which means that ``mobility'' is a construct consisting
of several variables. All variables but ``residential'' mobility are
relative changes of daily visitor numbers to the corresponding sectors
compared to the reference period. ``Residential'' mobility is the
relative change of daily time spent at residential areas.

@@

For different official and political measures we used one-hot encoded
daily variables, i.e.~ban of mass gatherings, school and kindergarten
closures and their gradual reopening, contact restrictions, and
mandatory face masks for shopping and public transport.

@@

\multicolumn{6}{l}{\textbf{Policies}}\\

@@

\hspace{1em}Age & constant & district & numeric & 2 variables: share of population $>=65$ years \& $<18$ years & INKAR \cite{inkar}\\

@@

All variables but the outcome ``Reported new cases of COVID-19'' and the
offset ``Active cases'' were centered for numerical stability. We did
not scale variables to unit variance to maintain interpretability of
effects on the original scale of variables. Additionally, we lagged the
effect of all variables (but outcome, offset, and the non-dynamic
socio-demographic variables) by 8 days (see Section Results) which
means that we assumed that their effects on the outcome will be visible
after 8 days.

@@

We used a directed acyclic graph as a graphical representation of the
hypothesized causal reasoning that leads to exposure to the SARS-CoV-2
virus, onset of COVID-19, and finally reports of COVID-19 cases. Every
node \(v_i\) in the graph is the graphical representation of an observed
or unobserved variable \(x_i\), a directed edge \(e_{ij}\) is an arrow
from node \(v_i\) to \(v_j\) that implies a direct causal relationship
from variable \(x_i\) onto variable \(x_j\). The set of all nodes is
denoted by \(V\), the set of all edges by \(E\), as such, the complete
DAG is the tuple \(G=(V,E)\). The seminal works of Spirtes and Pearl
\cite{spirtes2000causation, pearl2009causality} introduce the theory of
causal analysis, do-calculus, and how to analyze a DAG to estimate the
total or direct causal effect from a variable \(x_i\) onto a variable
\(x_j\). The direct effect is the effect associated with the edge
\(e_{ij}\) only (if it exists), while the total effect takes indirect
effects via other paths from \(v_i\) to \(v_j\) into account, too. Here
we estimated total effects only, since most of our variables were not
hypothesized to have a direct effect on the \emph{reported} number of
new COVID-19 cases. In contrast to prediction tasks, where one would
include all variables available, it is actually ill-advised to use all
available variables to estimate causal effects, due to introducing bias
by adjusting for unnecessary variables within the causal DAG. This is
why we need to identify a valid set of necessary variables (an
adjustment set) to estimate the proper causal effect
\cite{pearl2009causality}. The ``minimal adjustment set''
\cite{greenland1999causal} is a valid adjustment set of variables that
does not contain another valid adjustment set as a subset. However,
identifying a minimal adjustment set might not be enough to reliably
estimate the causal effect. Thus, we identified the ``optimal adjustment
set'' \cite{henckel2019graphical} as the set of variables which is a
valid adjustment set while having the lowest asymptotic variance in the
resulting causal effect estimates.

We analyzed the DAG from Fig~\ref{fig:dag-covid-19} with the R Software
\cite{rsoftware} and the R packages \texttt{dagitty} (formal
representation of the graph and minimal adjustment sets
\cite{textor_robust_2017}) and \texttt{pcalg} (for finding an optimal
adjustment set \cite{pcalg}). For the defined exposures and the outcome
``Reported new cases of COVID-19'', we computed the minimal and optimal
adjustment sets. Since it was possible that these sets contained
unobserved variables that needed to be left out of the regression model,
we chose the valid set with the highest pseudo-\(R^2\) (see next
section) to estimate the final total causal effect from exposure to
outcome.

@@

When we analyzed different adjustment sets given by analysis of the
causal DAG (i.e.~the minimal and optimal adjustment sets), we first
checked if the set included unobserved variables. If this was the case
for the optimal adjustment set, we discarded the unobserved variables
from the set and checked if it was still a valid adjustment set
(function \texttt{gac} in package \texttt{pcalg}
\cite{perkovi2015complete}). If a minimal adjustment set contained
unobserved variables, we discarded the whole set. We conducted a
log-linear regression (function \texttt{glm} with
\texttt{family=poisson()} for Poisson regression, and \texttt{glm.nb}
from the \texttt{MASS} package for the negative binomial regression
\cite{mass}) for every remaining valid adjustment set as regressors and
calculated a Pseudo-\(R^2\) given by \(1-V_m/V_0\), where \(V_m\) is the
sum of squared prediction errors of the current model and \(V_0\) is the
sum of squared prediction errors of the null model (intercept and offset
only). That is, our Pseudo-\(R^2\) is \(1\) minus the fraction of
variance unexplained. Finally, we decided for the model/adjustment set
with the highest pseudo-\(R^2\). We report the exponentiated estimated
coefficients along with 99 percent confidence intervals of the
estimates.

@@

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Descriptive statistics for observed variables.}}
%\fontsize{8}{10}\selectfont
\begin{tabular}[t]{ll}
\toprule
Variable & mean (SD)\\
\midrule
n & 54937\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Mobility}}\\
\hspace{1em}retail and recreation & -27.14 (24.59)\\
\hspace{1em}grocery and pharmacy & -4.04 (23.00)\\
\hspace{1em}parks & 46.33 (57.94)\\
\hspace{1em}workplaces & -22.98 (20.56)\\
\hspace{1em}residential & 8.20 (6.54)\\
\hspace{1em}transit stations & -29.83 (21.22)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Weather}}\\
\hspace{1em}Rainfall & 1.85 (3.98)\\
\hspace{1em}Temperature & 10.74 (5.27)\\
\hspace{1em}Humidity & 67.68 (13.10)\\
\hspace{1em}Wind & 3.64 (1.67)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Policies}}\\
\hspace{1em}Ban of mass gatherings & 0.82 (0.38)\\
\hspace{1em}School and kindergarten closures & 0.54 (0.36)\\
\hspace{1em}Contact restrictions & 0.74 (0.44)\\
\hspace{1em}Mandatory face masks & 0.48 (0.50)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Socio-demographic}}\\
\hspace{1em}Age (pop. 65 and older) & 22.09 (2.74)\\
\hspace{1em}Age (pop. younger 18) & 16.17 (1.25)\\
\hspace{1em}Gender & 0.51 (0.01)\\
\hspace{1em}Population density & 533.75 (701.84)\\
\hspace{1em}Foreign citizens & 10.03 (5.14)\\
\hspace{1em}Foreign citizens (refugees) & 1.88 (1.14)\\
\hspace{1em}Socio-economic status & 30.64 (6.02)\\
\hspace{1em}Turnout & 75.08 (3.79)\\
\hspace{1em}Right-wing populist party votes & 13.39 (5.32)\\
\hspace{1em}Nursing homes & 36.11 (30.69)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Awareness}}\\
\hspace{1em}Searches corona & 27.24 (18.31)\\
\hspace{1em}COVID-19 burden & 3.55 (10.37)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Case numbers (outcome and offset)}}\\
\hspace{1em}Reported new cases COVID-19 & 3.60 (10.39)\\
\hspace{1em}Active cases & 49.83 (121.96)\\
\bottomrule
\end{tabular}
\label{tab:descriptive-vars}
%\end{adjustwidth}
\end{table}

@@

In Germany, we observed a rebound in mobility after the initial
political measures, reductions in incident cases were associated with a
diminishing public interest in COVID-19, and temperatures were overall
increasing (cf.~Fig~\ref{fig:timeseries}); with correlations between
temporal progression and mobility in retail and recreation \(r_{A,D}=\)
-0.02, awareness (``Searches corona'') \(r_{A,C}=\) -0.29, and
temperature \(r_{A,B}=\) 0.79.

@@

We list the results of our causal analysis for the effects of our
variables in Table~\ref{tab:causaleffects}. The estimates are
multiplicative rates of increase/decrease for a one unit increase of the
respective variable: Values above 1 lead to an increase, below 1 to a
decrease of the infection rate. To put these estimates into perspective,
Fig~\ref{fig:effectsrange} shows the relative causal effect of the
different exposure variables on the number of reported COVID-19 cases on
a range of sensible values of the exposure variables (95 percent
quantiles of data points).

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Causal effect estimates with 99 percent confidence bands.}}
%\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lrrr}
\toprule
cause & effect estimate & CI (lower) & CI (upper)\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Mobility}}\\
\hspace{1em}retail and recreation & 1.0076 & 1.0056 & 1.0097\\
\hspace{1em}grocery and pharmacy & 0.9951 & 0.9938 & 0.9965\\
\hspace{1em}parks & 0.9998 & 0.9994 & 1.0003\\
\hspace{1em}transit stations & 0.9975 & 0.9946 & 1.0004\\
\hspace{1em}workplaces & 1.0031 & 1.0002 & 1.0059\\
\hspace{1em}residential & 1.0080 & 0.9987 & 1.0174\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Awareness}}\\
\hspace{1em}Searches corona & 1.0101 & 1.0085 & 1.0118\\
\hspace{1em}COVID-19 burden & 0.9942 & 0.9932 & 0.9951\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Weather}}\\
\hspace{1em}Temperature & 0.9918 & 0.9879 & 0.9958\\
\hspace{1em}Rainfall & 1.0108 & 1.0066 & 1.0150\\
\hspace{1em}Humidity & 1.0022 & 1.0006 & 1.0039\\
\hspace{1em}Wind & 0.9893 & 0.9802 & 0.9985\\
\bottomrule
\end{tabular}
\label{tab:causaleffects}
%\end{adjustwidth}
\end{table}

@@

Most research on determinants affecting case numbers of COVID-19 is
restricted to single aspects
\cite{fowler_effect_2020, li_retrospective_2020, shi_impact_2020, Wang2020temperature}.
To reliably identify causal drivers, one must adjust for confounders. To
this end, we used an integrated model with variables from different
aspects like mobility, awareness, weather, or socio-demographics and
identified confounders by causal analysis with a directed acyclic graph.
A causal approach is used in another current COVID-19 analysis
\cite{gencoglu2020causal}. There, however, they identify the causal
relationships (reconstruct a DAG), while we estimated causal effects for
a given hypothesized DAG.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on reported case numbers} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Edgar Steiger\textsuperscript{1*},
Tobias Mußgnug\textsuperscript{1},
Lars Eric Kroll\textsuperscript{1}%,
%Name4 Surname\textsuperscript{2},
%Name5 Surname\textsuperscript{2\ddag},
%Name6 Surname\textsuperscript{2\ddag},
%Name7 Surname\textsuperscript{1,2,3*},
%with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Central Research Institute of Ambulatory Health Care in Germany (Zi),
Salzufer 8, D-10587 Berlin, Germany
\\
%\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
%\\
%\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
%\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Several determinants are suspected to be causal drivers for new cases of COVID-19 infection. Correcting for possible confounders, we estimated the effects of the most prominent determining factors on reported case numbers. To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the determinants on new reported cases of COVID-19. Based on this, we computed valid adjustment sets of the possible confounding factors. We collected data for Germany from publicly available sources (e.g.~Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression. Our analysis revealed favorable effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy) or within residential areas, and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases. Conversely, we saw adverse effects leading to an increase in new COVID-19 cases for public mobility in retail and recreational areas or workplaces, awareness measured by searches for ``corona'' in Google, higher rainfall, and some socio-demographic factors. Non-pharmaceutical interventions were found to be effective in reducing case numbers. This comprehensive causal graph analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.

@@

Our objective is to provide valid estimates of the effects of the main drivers of the pandemic with a causal graph approach. We conducted a scoping review of the available studies regarding signaling pathways and determinants of the spread of SARS-CoV-2 infections and the reported new COVID-19 cases. Then we integrated the current findings into a directed acyclic graph for the progress of the pandemic at the regional level. Using the resulting model and the do-calculus we found identifiable effects without blocked causal paths whose effects can be analyzed with observational data \cite{Pearl_2014}. We used regional time series data of all German districts (\(401\)) from various publicly available sources to analyze these questions on a regional level. Germany is a good choice in this regard, because it has ample data on contributing factors on the regional level and has had high testing and treatment capacities from early on in the pandemic.

@@

When investigating causal determinants of SARS-CoV-2 infections, a
number of confounders have to be considered. Well-known risk factors for
SARS-CoV-2 as well as for other infections are demographic factors such
as age, gender, socio-economic status (SES), population density, and
foreign citizenship/ethnicity
\cite{de2020risk, Dragano2020.06.17.20133918, jhucovid19db2020}. In
Germany along with other countries (i.e.~Brazil, USA, or the UK),
populist parties or politicians and their electorate tend to be more
sceptical about effects of containment measures than the other part of
the electorate \cite{dohle_wingen_schreiber_2020, engle_staying_2020}.
Therefore we considered both ``right-wing populist party votes'' and
``voter turnout'' as possible confounders. Public health interventions
were also taken into account (contact restrictions, school closures
etc.), as their implementation showed strong correlations with
controlling the spread of SARS-CoV-2
\cite{cowling2020impact, juni_impact_2020, lai_effect_2020}. To avoid
bias due to reporting delay of case numbers we had to include weekday
and German holidays. We included some unobserved variables in our DAG
(e.g.~``Herd immunity''), too. Please note that ``Exposure to
SARS-CoV-2'' is itself an unobserved variable: German case numbers are
reported with delay after date of exposure and symptom onset.
\emph{Exposure to the virus} should not be confused with the formal
\emph{exposure variables} of the DAG.

@@

We collected and aggregated data on reported COVID-19 cases, regional
socio-demographic factors, weather, and general mobility on district and
state level in Germany for the period of 15 February 2020 to 8 July
2020. Our observation period for the outcome consisted of all dates from
20 February 2020 to 8 July 2020 (\(T=140\)), since we used a lag of 5
days for all confounders. We did not exclude any states or districts
(\(K=401\)). We analyzed the daily reported number of new cases as
outcome (\(K\cdot T=56{\ }140\) observations). The set of possible
predictors was derived from our causal DAG (see Table~\ref{tab:model-vars} and Fig~\ref{fig:dag-covid-19}). Due to
modelling and data limitations, some of the predictors were unobserved
or were modelled as a construct consisting of several variables. For our causal graph analysis of effects, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).

@@

To assess the mobility of the German population, we used data publicly
available on German state level from Google \cite{google_mobility}.
Measurements are daily relative changes of mobility in percent compared
to the period of 3 January 2020 to 6 February 2020. Missing values
(\(25\) out of \(13{\ }488\)) were imputed with value \(0\) and the
state level measurements were passed onto districts within the
corresponding state. Google mobility data was available for six
different sectors of daily life (``retail and recreation'', ``grocery
and pharmacy'', ``parks'', ``transit stations'', ``workplaces'',
``residential'') which means that ``mobility'' is a construct consisting
of several variables. All variables but ``residential'' mobility are
relative changes of daily visitor numbers to the corresponding sectors
compared to the reference period. ``Residential'' mobility is the
relative change of daily time spent at residential areas. The six mobility variables showed high correlations among each other and with other variables. To reduce multicollinearity, we transformed them by principal component analysis (PCA) into six uncorrelated principal components which were used in place of the original variables.

@@

For different official and political interventions on a day and district level we used one-hot encoded daily variables, i.e.~ban of mass gatherings, school and kindergarten closures and their gradual reopening, contact restrictions, and mandatory face masks for shopping and public transport.

@@

\multicolumn{6}{l}{\textbf{Interventions}}\\

@@

\hspace{1em}Age & constant & district & numeric & 2 variables: share of population >=65 years \& <18 years & INKAR \cite{inkar}\\

@@

All continuous variables but the outcome ``Reported new cases of COVID-19'' and the offset ``Active cases'' were centered and scaled by one standard deviation for numerical stability, while we left binary variables as-is. After estimating the effects of variables, we re-scaled continuous variables' effects to their original scale. Additionally for mobility variables, we re-transformed the effects of the principal components to the original mobility variables. Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (see Section Results) which means that we assumed that their effects on the outcome will be visible after 5 days.

@@

We used a directed acyclic graph as a graphical representation of the
hypothesized causal reasoning that leads to exposure to the SARS-CoV-2
virus, onset of COVID-19, and finally reports of COVID-19 cases. We use the terms ``causal effect'' or ``causal relationship'' for effect estimates that are based on this causal graph framework. Every
node \(v_i\) in the graph is the graphical representation of an observed
or unobserved variable \(x_i\), a directed edge \(e_{ij}\) is an arrow
from node \(v_i\) to \(v_j\) that implies a direct causal relationship
from variable \(x_i\) onto variable \(x_j\). The set of all nodes is
denoted by \(V\), the set of all edges by \(E\), as such, the complete
DAG is the tuple \(G=(V,E)\). The seminal works of Spirtes and Pearl
\cite{spirtes2000causation, pearl2009causality} introduce the theory of
causal analysis, do-calculus, and how to analyze a DAG to estimate the
total or direct causal effect from a variable \(x_i\) onto a variable
\(x_j\). The direct effect is the effect associated with the edge
\(e_{ij}\) only (if it exists), while the total effect takes indirect
effects via other paths from \(v_i\) to \(v_j\) into account, too. Here
we estimated total effects only, since most of our variables were not
hypothesized to have a direct effect on the \emph{reported} number of
new COVID-19 cases. In contrast to prediction tasks, where one would
include all variables available, it is actually ill-advised to use all
available variables to estimate causal effects, due to introducing bias
by adjusting for unnecessary variables within the causal DAG. This is
why we need to identify a valid set of necessary variables (an
adjustment set) to estimate the proper causal effect
\cite{pearl2009causality}. The ``minimal adjustment set''
\cite{greenland1999causal} is a valid adjustment set of variables that
does not contain another valid adjustment set as a subset. However,
identifying a minimal adjustment set might not be enough to reliably
estimate the causal effect. Thus, we identified the ``optimal adjustment
set'' \cite{henckel2019graphical} as the set of variables which is a
valid adjustment set while having the lowest AIC.

We analyzed the DAG from Fig~\ref{fig:dag-covid-19} with the R Software
\cite{rsoftware} and the R packages \texttt{dagitty} (formal
representation of the graph and minimal adjustment sets
\cite{textor_robust_2017}) and \texttt{pcalg} (for finding an optimal
adjustment set \cite{pcalg}). For the defined exposures and the outcome
``Reported new cases of COVID-19'', we computed the minimal and optimal
adjustment sets. Since it was possible that these sets contained unobserved variables that needed to be left out of the regression model, we chose the valid set with the lowest AIC (see next section) to estimate the final total causal effect from exposure to outcome.

@@

We conducted a log-linear regression (function \texttt{glm} with \texttt{family=poisson()} for Poisson regression, and \texttt{glm.nb} from the \texttt{MASS} package for the negative binomial regression \cite{mass}) for the full data set to assess general model adequacy and to estimate the \(\theta\) parameter of the negative binomial. The proper lag between exposures and outcome was found by 10-fold cross-validation on different lags between 1 and 20 days. Model diagnostics on the final full model did not show severe problems with model assumptions (linearity, distribution of residuals, independence of observations). Analysis of variance inflation factors revealed some problems with multicollinearity. To reduce the effects of multicollinearity, first we transformed the highly correlated mobility variables by PCA as described above. Second, we used a ridge regression approach \cite{ridge1970}, which is a regularization method that shrinks regression coefficients and alleviates the effect of correlation between variables on their respective regression coefficients. Furthermore, regularized regression allows for better fits on unseen data, thus preventing overfitting the data, too. The hyper-parameter \(\lambda\) of the ridge regression was chosen by 10-fold cross-validation, where the folds were constructed from random subsets of the \(401\) districts. We used this hyper-parameter with the \texttt{cv.glmnet} function from the R package \texttt{glmnet} \cite{glmnet2010} with \texttt{family=negative.binomial(theta)} and chose the \(\lambda\) value within one standard deviation from the minimal \(\lambda\) as regularization hyper-parameter. Afterwards, we calculated the effects of separate exposures on the outcome. For every exposure, we analyzed the different valid adjustment sets given by analysis of the causal DAG (i.e.~the minimal and optimal adjustment sets). Then, we first checked if the respective set included unobserved variables. If this was the case for the optimal adjustment set, we discarded the unobserved variables from the set and checked if it was still a valid adjustment set (function \texttt{gac} in package \texttt{pcalg} \cite{perkovi2015complete}). If a minimal adjustment set contained unobserved variables, we discarded the whole set. If no valid adjustment set for a given exposure was available, we concluded that the effect of this exposure was unidentifiable within our causal graph. We used the function \texttt{glmnet} with the parameters \(\theta\) and \(\lambda\) as above on every remaining valid adjustment set as regressors (that is, we applied ridge regression) and calculated the Akaike information criterion (AIC) for this model/set of regressors. Finally, for every exposure, we decided for the model/adjustment set (if available) with the lowest AIC. We report the exponentiated estimated coefficients for the separate exposures on their original scale.

@@

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Descriptive statistics for observed variables.}}
%\fontsize{8}{10}\selectfont
\begin{tabular}[t]{ll}
\toprule
Variable & mean (SD)\\
\midrule
n & 56140\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Mobility}}\\
\hspace{1em}retail and recreation & -26.62 (24.60)\\
\hspace{1em}grocery and pharmacy & -3.94 (22.77)\\
\hspace{1em}parks & 47.26 (58.20)\\
\hspace{1em}workplaces & -22.96 (20.35)\\
\hspace{1em}residential & 8.13 (6.49)\\
\hspace{1em}transit stations & -29.58 (21.11)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Awareness}}\\
\hspace{1em}Searches corona & 26.94 (18.23)\\
\hspace{1em}COVID-19 burden & 3.50 (10.28)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Weather}}\\
\hspace{1em}Rainfall & 1.89 (4.01)\\
\hspace{1em}Temperature & 10.90 (5.33)\\
\hspace{1em}Humidity & 67.81 (13.03)\\
\hspace{1em}Wind & 3.63 (1.66)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Interventions}}\\
\hspace{1em}Ban of mass gatherings & 0.83 (0.38)\\
\hspace{1em}School and kindergarten closures & 0.54 (0.36)\\
\hspace{1em}Contact restrictions & 0.74 (0.44)\\
\hspace{1em}Mandatory face masks & 0.49 (0.50)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Socio-demographic}}\\
\hspace{1em}Age (pop. 65 and older) & 22.09 (2.74)\\
\hspace{1em}Age (pop. younger 18) & 16.17 (1.25)\\
\hspace{1em}Gender & 50.59 (0.64)\\
\hspace{1em}Population density & 533.75 (701.84)\\
\hspace{1em}Foreign citizens & 10.03 (5.14)\\
\hspace{1em}Foreign citizens (refugees) & 1.88 (1.14)\\
\hspace{1em}Socio-economic status & 30.64 (6.02)\\
\hspace{1em}Turnout & 75.08 (3.79)\\
\hspace{1em}Right-wing populist party votes & 13.39 (5.32)\\
\hspace{1em}Nursing homes & 36.11 (30.69)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Case numbers (Outcome and offset)}}\\
\hspace{1em}Reported new cases COVID-19 & 3.53 (10.29)\\
\hspace{1em}Active cases & 48.76 (120.86)\\
\bottomrule
\end{tabular}
\label{tab:descriptive-vars}
%\end{adjustwidth}
\end{table}

@@

In Germany, we observed a rebound in mobility after the initial political measures, reductions in incident cases were associated with a diminishing public interest in COVID-19, and temperatures were overall increasing (cf.~Fig~\ref{fig:timeseries}); with correlations between temporal progression and mobility in retail and recreation \(r_{A,B}=\) 0.02, awareness (``Searches corona'') \(r_{A,C}=\) -0.3, and temperature \(r_{A,D}=\) 0.8.

@@

We list the results of our causal analysis for the effects of different exposure variables in Table~\ref{tab:causaleffects}. The estimates are multiplicative rates of increase/decrease for a one unit increase of the respective variable: Values above 1 lead to an increase, below 1 to a decrease of the infection rate. To put these estimates into perspective, Fig~\ref{fig:effectsrange} shows the relative causal effect of the different exposure variables on the number of reported COVID-19 cases on a range of sensible values of the exposure variables (95 percent quantiles of data points).

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Effect estimates from causal graph analysis.}}
%\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lr}
\toprule
cause & effect estimate\\
\midrule
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Mobility}}\\
\hspace{1em}retail and recreation & 1.0011\\
\hspace{1em}grocery and pharmacy & 0.9977\\
\hspace{1em}parks & 0.9997\\
\hspace{1em}transit stations & 1.0026\\
\hspace{1em}workplaces & 1.0033\\
\hspace{1em}residential & 0.9903\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Awareness}}\\
\hspace{1em}Searches corona & 1.0089\\
\hspace{1em}COVID-19 burden & 0.9980\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Weather}}\\
\hspace{1em}Temperature & 0.9905\\
\hspace{1em}Rainfall & 1.0121\\
\hspace{1em}Humidity & 1.0057\\
\hspace{1em}Wind & 1.0329\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Interventions}}\\
\hspace{1em}Interventions (ban of mass gatherings) & 0.9729\\
\hspace{1em}Interventions (school and kindergarten closures) & 0.9277\\
\hspace{1em}Interventions (contact restrictions) & 0.8314\\
\hspace{1em}Interventions (mandatory face masks) & 0.9064\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Demographic}}\\
\hspace{1em}Age (pop. 65 and older) & 0.9953\\
\hspace{1em}Age (pop. younger 18) & 1.0120\\
\hspace{1em}Foreign citizens & 1.0048\\
\hspace{1em}Foreign citizens (refugees) & 0.9985\\
\hspace{1em}Gender & 0.9925\\
\hspace{1em}Nursing homes & 1.0011\\
\hspace{1em}Population density & 1.0000\\
\hspace{1em}Socio-economic status & 0.9982\\
\bottomrule
\end{tabular}
\label{tab:causaleffects}
%\end{adjustwidth}
\end{table}

@@

Most research on determinants affecting case numbers of COVID-19 is restricted to single aspects \cite{fowler_effect_2020, li_retrospective_2020, shi_impact_2020, Wang2020temperature}. To reliably identify (causal) drivers, one must adjust for confounders. To this end, we used an integrated model with variables from different aspects like mobility, awareness, weather, or socio-demographics and identified confounders by causal analysis with a directed acyclic graph. A causal approach is used in another current COVID-19 analysis \cite{gencoglu2020causal}. There, however, they identify the causal relationships (reconstruct a DAG), while we estimated effects for a given hypothesized causal DAG.
